HMMotion: Using tracking data to predict coverage strategies in American Football

### 2025-01-06 submission.Rmd
.

We aim to forecast the defensive scheme (man- or zone defense) using the `pff_passCoverage` indicator in play-by play data. We omit plays tagged as \textit{others} as well as plays with more than five offensive linemen and with two quarterbacks. Since we are specifically interested in analyzing pre-snap player movements, we concentrate on plays that contain any pre-snap motion. Ultimately, we end up with $3985$ offensive plays in total, from which the defense played $2973$ in zone and $1012$ in man coverage.

@@

We compare the predictive performance of our three models to determine the effectiveness of player movements before snap for detecting the correct defensive scheme. We train our the models on 85% of the data and evaluate performance on the remaining 15%. Since we use cross-validation for tuning of the models during the training phase, we perform our evaluation repeatedly. We use 50 repetition each time splitting the data differently, and for each we tune and train a model and then evaluate the performance on the test set. We do this to ensure that our results are meaningful and not due to chance (especially in our small data set).

@@

As frequently mentioned, we face the problem of only having a small amount of relevant data available. In order to avoid overfitting problems, we therefore focus on a basic feature set of 32 variables for our main results: 6 convex hull related features (3 for offense and defense, respectively), 20 player positions features (10 standardized $x$ and $y$ coordinates, 5 for offense, 5 for defense), and 6 play-by-play features. However, using more thoroughly crafted features such as distances and orientation as described above, we can enlarge the feature set to 67 total variables: 30 distance variables (for each of the 10 relevant players the total distance, $x$-distances, and $y$-distances to the football), and 5 orientation variables (for each defender the orientation with respect to the QB).

In the following, we provide a comparison of these enlarged models and the smaller ones used for our main results. When comparing results of a binary outcome variable, accuracy can be a misleading metric, since it is dependent on the threshold selected for classification of the outcome. Usually, for probabilistic predictions as obtained by the two model classes considered, one uses the naive threshold of 0.5. However doing so is arbitrary and especially for imbalanced data, changing the threshold may change the results of the accuracy drastically. AUC (or area under the ROC curve) on the other hand, tries to evaluated the performance of a model over a suitable grid of thresholds, and is thus avoids the subjective choice of threshold. Another popularly used metric is the logloss (or negative log-likelihood loss), which is often a preferred choice due to being a proper scoring rule (mathematically speaking a scoring rule is proper, if it is optimal in expectation under the true probability). While using the logloss as evaluation metric is mathematically the best option, interpretation is not as intuitive. Specifically, whether a logloss of a model can be considered good, depends on the (im)balance of the classes (as well as the number of classes, which in a our binary case is only 2). We will therefore focus on these two metrics for the evaluation of the models.

>>>>

We aim to forecast the defensive scheme (man- or zone defense) using the `pff_passCoverage` indicator in play-by play data. We omit plays tagged as *others* as well as plays with more than five offensive linemen and with two quarterbacks. Since we are specifically interested in analyzing pre-snap player movements, we concentrate on plays that contain any pre-snap motion. Ultimately, we end up with $3985$ offensive plays in total, from which the defense played $2973$ in zone and $1012$ in man coverage.

@@

We compare the predictive performance of our three models to determine the effectiveness of player movements before snap for detecting the correct defensive scheme. We train the models on 85% of the data and evaluate performance on the remaining 15%. Since we use cross-validation for tuning of the models during the training phase, we perform our evaluation repeatedly. We use 50 repetition each time splitting the data differently, and for each we tune and train a model and then evaluate the performance on the test set. We do this to ensure that our results are meaningful and not due to chance (especially in our small data set).

@@

As frequently mentioned, we face the problem of only having a small amount of relevant data available. In order to avoid overfitting problems, we therefore focus on a basic feature set of 32 variables for our main results: 6 convex hull related features (3 for offense and defense, respectively), 20 player positions features (10 standardized $x$ and 10 standardized $y$ coordinates, 5 for offense, 5 for defense), and 6 play-by-play features. However, using more thoroughly crafted features such as distances and orientation as described above, we can enlarge the feature set to 67 total variables: 30 distance variables (for each of the 10 relevant players the total distance, $x$-distances, and $y$-distances to the football), and 5 orientation variables (for each defender the orientation with respect to the QB).

In the following, we provide a comparison of these enlarged models and the smaller ones used for our main results. When comparing results of a binary outcome variable, accuracy can be a misleading metric, since it is dependent on the threshold selected for classification of the outcome. Usually, for probabilistic predictions as obtained by the two model classes considered, one uses the naive threshold of 0.5. However doing so is arbitrary and especially for imbalanced data, changing the threshold may change the results of the accuracy drastically. AUC (or area under the ROC curve) on the other hand, tries to evaluated the performance of a model over a suitable grid of thresholds, and thus avoids the subjective choice of threshold. Another popularly used metric is the logloss (or negative log-likelihood loss), which is often a preferred choice due to being a proper scoring rule. While using the logloss as evaluation metric is mathematically the best option, interpretation is not as intuitive. Specifically, whether a logloss of a model can be considered good, depends on the (im)balance of the classes (as well as the number of classes, which in a our binary case is only 2). Without going into further detail, we therefore focus on these two metrics for the evaluation of the models.
